---
title: "Secretome-Aware Hallmark Pathway Scoring and Correlation"
author:
  - name: Yuhan Sun
    email: yzsun0123@gmail.com
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Secretome-Aware Hallmark Pathway Scoring and Correlation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = FALSE
)
```

# Motivation

When hallmark pathway scores are correlated against expression of a gene
set that itself contributes genes to those hallmarks (for example the
**secretome**, many of whose genes appear in MSigDB Hallmark
collections), the correlation is inflated: the score depends in part on
the very genes being correlated against it. This is a circular-analysis
problem, not a biological signal.

`r Biocpkg("surVisualiser")` addresses this by:

1. Removing the confounding gene list from the expression matrix
   **before** ranking and running GSEA, so the resulting enrichment
   scores cannot share genes with the gene list.
2. Correlating the cleaned hallmark scores against expression of the
   same gene list, producing a Pearson correlation table.
3. Rendering a three-panel (Tumor / Normal / Tumor - Normal) heatmap of
   the correlations for a given cancer type.

This vignette walks through the full workflow end to end.

# Installation

```{r install}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("surVisualiser")

# Suggested packages used by this workflow:
BiocManager::install(c(
    "msigdbr", "clusterProfiler",
    "tidyHeatmap", "ComplexHeatmap",
    "circlize", "RColorBrewer"
))
```

# Setup

```{r load}
library(surVisualiser)
data(secretome_genes, package = "surVisualiser")
head(secretome_genes)
```

The bundled `secretome_genes` data frame provides the curated secretome
gene list. Its `symbol` column supplies gene symbols suitable for the
`excluded_genes` argument to `gsea_test_scrto_gene()`.

# Step 1: Per-sample Pre-ranked GSEA with Secretome Exclusion

Assume `expr_tumor` and `expr_normal` are tibbles with columns
`sample`, `symbol`, and `raw_count_scaled_adjusted` (as produced by
`TCGA_transcript()` with scaling). Run the GSEA-based hallmark scoring
separately for tumor and normal samples, removing secretome genes from
the expression input before ranking:

```{r gsea}
hallmarks_tumor <- gsea_test_scrto_gene(
    data              = expr_tumor,
    rank_by           = "raw_count_scaled_adjusted",
    excluded_genes    = secretome_genes$symbol,
    species           = "Homo sapiens",
    gene_collections  = "h"
)

hallmarks_normal <- gsea_test_scrto_gene(
    data              = expr_normal,
    rank_by           = "raw_count_scaled_adjusted",
    excluded_genes    = secretome_genes$symbol,
    gene_collections  = "h"
)
```

Each call returns a long tibble with one row per
(sample, hallmark) pair, including the normalised enrichment score
(`NES`), `pvalue`, `p.adjust`, and other `clusterProfiler::GSEA` output
columns.

**Note.** `gsea_test_scrto_gene()` performs per-sample pre-ranked GSEA
and returns an enrichment statistic per pathway per sample. It is
mathematically distinct from matrix-based per-sample scoring methods
such as GSVA or ssGSEA. For correlation-oriented workflows either
family of methods is valid; GSEA is used here to preserve parity with
the historical `surVisualiser` analysis pipeline.

# Step 2: Pearson Correlation Against Secretome Expression

Pass the hallmark scores and the expression table into
`corr_pathway_scrto()` to compute Pearson correlations between every
hallmark and every secretome gene:

```{r corr}
corr_tumor <- corr_pathway_scrto(
    hallmark_scores    = hallmarks_tumor,
    expression         = expr_tumor,
    secretome_symbols  = secretome_genes$symbol
)

corr_normal <- corr_pathway_scrto(
    hallmark_scores    = hallmarks_normal,
    expression         = expr_normal,
    secretome_symbols  = secretome_genes$symbol
)

head(corr_tumor)
```

Both outputs are tidy long tibbles with columns `pathway`, `gene`,
`value` (Pearson r), and `p.value`. Because secretome genes were
excluded from the GSEA input in Step 1, the correlations on both sides
do not share genes and can be interpreted without the circular-analysis
concern.

# Step 3: Three-Panel Tumor / Normal / Difference Heatmap

`three_heatmaps_of_pth_scrto_TN()` renders a single PDF with three
side-by-side `tidyHeatmap` panels: the tumor correlations, the normal
correlations, and their element-wise difference. The output directory
is a function argument.

```{r heatmap}
result <- three_heatmaps_of_pth_scrto_TN(
    tumor_corr   = corr_tumor,
    normal_corr  = corr_normal,
    cancer       = "BRCA",
    output_dir   = "figures/BRCA",
    p_cutoff     = 0.05
)

result$file
```

Non-significant correlations are set to zero before the difference is
computed. The returned list also carries the three `tidyHeatmap`
objects under `$tumor`, `$normal`, and `$diff` for further assembly or
inspection.

# Session Info

```{r session-info, eval=TRUE}
sessionInfo()
```
