---
title: "Introduction to surVisualiser"
author:
  - name: Yuhan Sun
    email: yzsun0123@gmail.com
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Introduction to surVisualiser}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Introduction

**surVisualiser** provides a streamlined workflow for analysing TCGA
transcriptome data. The package covers the full analysis pipeline from
raw transcript counts to publication-ready survival plots:
1. Reading and aggregating TCGA transcript-level expression data
2. Integrating clinical survival metadata
3. Immune cell deconvolution via CIBERSORT
4. Kaplan-Meier survival visualisation with flexible grouping

# Installation

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("surVisualiser")
```

# Quick Start

## Step 1: Load TCGA Expression Data

```{r load-data, eval=FALSE}
library(surVisualiser)

# Read tumor samples for breast cancer
brca <- TCGA_transcript(
    cancer = "BRCA",
    data_dir = "/path/to/TCGA/data",
    sample_type = "tumor"
)
```

## Step 2: Combine with Clinical Data

```{r clinical, eval=FALSE}
brca_clinical <- clinical_combine(
    brca,
    cancer = "BRCA",
    clinical_dir = "/path/to/clinical/data"
)
```

## Step 3: Immune Cell Deconvolution

```{r cibersort, eval=FALSE}
# Load bundled reference matrix
data(my_ref)

deconv <- CIBERSORT(
    brca,
    c1 = "sample",
    c2 = "symbol",
    c3 = "raw_count",
    ref = my_ref,
    meth = "cibersort",
    act = "get"
)
```

## Step 4: Survival Analysis

```{r survival, eval=FALSE}
# Split gene expression at median
surv_data <- split_median(brca_clinical,
    cat = "cancer_stage",
    item = "TP53"
)

# Generate Kaplan-Meier plot
ggsurvp_cat_item(
    surv_data,
    p = TRUE, nrow = 2,
    palette = c("#E41A1C", "#377EB8"),
    xlab = "Days", ylab = "Survival Probability",
    title = "TP53 Expression and Survival",
    size0 = 12
)
```

# Bundled Data

The package includes three datasets:

- **`cellname`**: Mapping from CIBERSORT cell-type IDs to readable names
- **`secretome_genes`**: Comprehensive list of secreted protein genes
- **`my_ref`**: Reference matrix for CIBERSORT deconvolution

```{r data-example}
data(cellname, package = "surVisualiser")
head(cellname)
```

# Session Info

```{r session-info}
sessionInfo()
```
